---
title: "Predicting Loan Repayment"
author: "Quang Duong"
date: "July 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In the lending industry, investors provide loans to borrowers in exchange for the promise of repayment with interest. If the borrower repays the loan, then the lender profits from the interest. However, if the borrower is unable to repay the loan, then the lender loses money. Therefore, lenders face the problem of predicting the risk of a borrower being unable to repay a loan.

To address this problem, we will use publicly available data from LendingClub.com, a website that connects borrowers and investors over the Internet. This dataset represents 9,578 3-year loans that were funded through the LendingClub.com platform between May 2007 and February 2010. The binary dependent variable not_fully_paid indicates that the loan was not paid back in full (the borrower either defaulted or the loan was "charged off," meaning the borrower was deemed unlikely to ever pay it back).

To predict this dependent variable, we will use the following independent variables available to the investor when deciding whether to fund a loan:

credit.policy: 1 if the customer meets the credit underwriting criteria of LendingClub.com, and 0 otherwise.
purpose: The purpose of the loan (takes values "credit_card", "debt_consolidation", "educational", "major_purchase", "small_business", and "all_other").
int.rate: The interest rate of the loan, as a proportion (a rate of 11% would be stored as 0.11). Borrowers judged by LendingClub.com to be more risky are assigned higher interest rates.
installment: The monthly installments ($) owed by the borrower if the loan is funded.
log.annual.inc: The natural log of the self-reported annual income of the borrower.
dti: The debt-to-income ratio of the borrower (amount of debt divided by annual income).
fico: The FICO credit score of the borrower.
days.with.cr.line: The number of days the borrower has had a credit line.
revol.bal: The borrower's revolving balance (amount unpaid at the end of the credit card billing cycle).
revol.util: The borrower's revolving line utilization rate (the amount of the credit line used relative to total credit available).
inq.last.6mths: The borrower's number of inquiries by creditors in the last 6 months.
delinq.2yrs: The number of times the borrower had been 30+ days past due on a payment in the past 2 years.
pub.rec: The borrower's number of derogatory public records (bankruptcy filings, tax liens, or judgments).

## Preparing the Dataset

Load the dataset loans.csv into a data frame called loans, and explore it using the str() and summary() functions.

What proportion of the loans in the dataset were not paid in full? Please input a number between 0 and 1.

```{r results='hold'}
loans <- read.csv('loans.csv')
attach(loans)
table(loans$not.fully.paid)
1533/(1533+8045)
```

Which of the following variables has at least one missing observation?

```{r}
summary(loans)
```

Which of the following is the best reason to fill in the missing values for these variables instead of removing observations with missing data? (Hint: you can use the subset() function to build a data frame with the observations missing at least one value. To test if a variable, for example pub.rec, is missing a value, use is.na(pub.rec).)

```{r}
missing <- subset(loans, is.na(log.annual.inc) | is.na(days.with.cr.line) | is.na(revol.util) | is.na(inq.last.6mths) | is.na(delinq.2yrs) | is.na(pub.rec))
table(missing$not.fully.paid)
12/(50+12)
```

We see that only 62 of 9578 loans have missing data; removing this small number of observations would not lead to overfitting. From table(missing$not.fully.paid), we see that 12 of 62 loans with missing data were not fully paid, or 19.35%. This rate is similar to the 16.01% across all loans, so the form of biasing described is not an issue. However, to predict risk for loans with missing data we need to fill in the missing values instead of removing the observations.

For the rest of this problem, we'll be using a revised version of the dataset that has the missing values filled in with multiple imputation (which was discussed in the Recitation of this Unit). To ensure everybody has the same data frame going forward, you can either run the commands below in your R console (if you haven't already, run the command install.packages("mice") first), or you can download and load into R the dataset we created after running the imputation: loans_imputed.csv.

```{r results='hide'}
# library(mice)
# set.seed(144)
# vars.for.imputation = setdiff(names(loans), "not.fully.paid")
# imputed = complete(mice(loans[vars.for.imputation]))
# loans[vars.for.imputation] = imputed
loans <- read.csv('loans_imputed.csv')
```

## Prediction Models

Now that we have prepared the dataset, we need to split it into a training and testing set. To ensure everybody obtains the same split, set the random seed to 144 (even though you already did so earlier in the problem) and use the sample.split function to select the 70% of observations for the training set (the dependent variable for sample.split is not.fully.paid). Name the data frames train and test.

Now, use logistic regression trained on the training set to predict the dependent variable not.fully.paid using all the independent variables.

Which independent variables are significant in our model?

```{r}
library(caTools)
set.seed(144)
split <- sample.split(loans$not.fully.paid, SplitRatio = 0.7)
train <- subset(loans, split==TRUE)
test <- subset(loans, split==FALSE)
notPaidLog <- glm(not.fully.paid ~., data = train, family = binomial)
summary(notPaidLog)
```

Consider two loan applications, which are identical other than the fact that the borrower in Application A has FICO credit score 700 while the borrower in Application B has FICO credit score 710.

Let Logit(A) be the log odds of loan A not being paid back in full, according to our logistic regression model, and define Logit(B) similarly for loan B. What is the value of Logit(A) - Logit(B)?

```{r}
-9.406e-03*(700-710)
exp(0.09406)
```

Predict the probability of the test set loans not being paid back in full (remember type="response" for the predict function). Store these predicted probabilities in a variable named predicted.risk and add it to your test set (we will use this variable in later parts of the problem). Compute the confusion matrix using a threshold of 0.5.

What is the accuracy of the logistic regression model? Input the accuracy as a number between 0 and 1.

```{r}
predicted.risk <- predict(notPaidLog, newdata = test, type = "response")
test$predicted.risk <- predicted.risk
table(test$not.fully.paid, test$predicted.risk>=0.5)
(2400+3)/(2400+13+457+3)
(2400+13)/(2400+13+457+3)
```

Use the ROCR package to compute the test set AUC.

```{r}
library(ROCR)
predROCR <- prediction(test$predicted.risk, test$not.fully.paid)
as.numeric(performance(predROCR, "auc")@y.values)
```

## A "Smart Baseline"

In the previous problem, we built a logistic regression model that has an AUC significantly higher than the AUC of 0.5 that would be obtained by randomly ordering observations.

However, LendingClub.com assigns the interest rate to a loan based on their estimate of that loan's risk. This variable, int.rate, is an independent variable in our dataset. In this part, we will investigate using the loan's interest rate as a "smart baseline" to order the loans according to risk.

Using the training set, build a bivariate logistic regression model (aka a logistic regression model with a single independent variable) that predicts the dependent variable not.fully.paid using only the variable int.rate.

The variable int.rate is highly significant in the bivariate model, but it is not significant at the 0.05 level in the model trained with all the independent variables. What is the most likely explanation for this difference?

```{r}
bivariateLog <- glm(not.fully.paid ~ int.rate, data = train, family = binomial)
summary(bivariateLog)
```

int.rate is correlated with other risk-related variables, and therefore does not incrementally improve the model when those other variables are included.

Make test set predictions for the bivariate model. What is the highest predicted probability of a loan not being paid in full on the testing set?

```{r}
bivariatePred <- predict(bivariateLog, newdata = test, type = "response")
max(bivariatePred)
```

With a logistic regression cutoff of 0.5, how many loans would be predicted as not being paid in full on the testing set?

```{r}
summary(bivariatePred)
```

According to the summary function, the maximum predicted probability of the loan not being paid back is 0.4266, which means no loans would be flagged at a logistic regression cutoff of 0.5.

What is the test set AUC of the bivariate model?

```{r}
bivaPredROCR <- prediction(bivariatePred, test$not.fully.paid)
as.numeric(performance(bivaPredROCR,"auc")@y.values)
```

## Computing the Profitability of an Investment

While thus far we have predicted if a loan will be paid back or not, an investor needs to identify loans that are expected to be profitable. If the loan is paid back in full, then the investor makes interest on the loan. However, if the loan is not paid back, the investor loses the money invested. Therefore, the investor should seek loans that best balance this risk and reward.

To compute interest revenue, consider a $c investment in a loan that has an annual interest rate r over a period of t years. Using continuous compounding of interest, this investment pays back c * exp(rt) dollars by the end of the t years, where exp(rt) is e raised to the r*t power.

How much does a $10 investment with an annual interest rate of 6% pay back after 3 years, using continuous compounding of interest? Hint: remember to convert the percentage to a proportion before doing the math. Enter the number of dollars, without the $ sign.

```{r}
10*exp(0.06*3)
```

While the investment has value c * exp(rt) dollars after collecting interest, the investor had to pay $c for the investment. What is the profit to the investor if the investment is paid back in full?

c * exp(rt) - c correct

Now, consider the case where the investor made a $c investment, but it was not paid back in full. Assume, conservatively, that no money was received from the borrower (often a lender will receive some but not all of the value of the loan, making this a pessimistic assumption of how much is received). What is the profit to the investor in this scenario?

-c 

In the previous subproblem, we concluded that an investor who invested c dollars in a loan with interest rate r for t years makes c * (exp(rt) - 1) dollars of profit if the loan is paid back in full and -c dollars of profit if the loan is not paid back in full (pessimistically).

In order to evaluate the quality of an investment strategy, we need to compute this profit for each loan in the test set. For this variable, we will assume a $1 investment (aka c=1). To create the variable, we first assign to the profit for a fully paid loan, exp(rt)-1, to every observation, and we then replace this value with -1 in the cases where the loan was not paid in full. All the loans in our dataset are 3-year loans, meaning t=3 in our calculations. Enter the following commands in your R console to create this new variable:

```{r}
test$profit <- exp(test$int.rate*3)-1
test$profit[test$not.fully.paid==1] = -1
```

What is the maximum profit of a $10 investment in any loan in the testing set (do not include the $ sign in your answer)?

```{r}
10*max(test$profit)
```

## An Investment Strategy Based on Risk

A simple investment strategy of equally investing in all the loans would yield profit $20.94 for a $100 investment. But this simple investment strategy does not leverage the prediction model we built earlier in this problem. As stated earlier, investors seek loans that balance reward with risk, in that they simultaneously have high interest rates and a low risk of not being paid back.

To meet this objective, we will analyze an investment strategy in which the investor only purchases loans with a high interest rate (a rate of at least 15%), but amongst these loans selects the ones with the lowest predicted risk of not being paid back in full. We will model an investor who invests $1 in each of the most promising 100 loans.

First, use the subset() function to build a data frame called highInterest consisting of the test set loans with an interest rate of at least 15%.

What is the average profit of a $1 investment in one of these high-interest loans (do not include the $ sign in your answer)?

```{r}
highInterest  <- subset(test, test$int.rate>=0.15)
mean(highInterest$profit)
table(highInterest$not.fully.paid)
110/(110+327)
```

Next, we will determine the 100th smallest predicted probability of not paying in full by sorting the predicted risks in increasing order and selecting the 100th element of this sorted list. Find the highest predicted risk that we will include by typing the following command into your R console:

```{r}
cuttoff <- sort(highInterest$predicted.risk, decreasing = FALSE)[100]
```

Use the subset() function to build a data frame called selectedLoans consisting of the high-interest loans with predicted risk not exceeding the cutoff we just computed. Check to make sure you have selected 100 loans for investment.

```{r}
selectedLoans <- subset(highInterest, highInterest$predicted.risk <= cuttoff)
str(selectedLoans)
```

What is the profit of the investor, who invested $1 in each of these 100 loans (do not include the $ sign in your answer)?

```{r}
sum(selectedLoans$profit)
```

How many of 100 selected loans were not paid back in full?

```{r}
table(selectedLoans$not.fully.paid)
```

