---
title: "Visualizing Attributes of Parole Violators "
author: "Quang Duong"
date: "`r Sys.Date()`"
output:
    # pdf_document:
    #   highlight: pygments
    #   fig_width: 4
    #   fig_height: 3
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, results = "hold", message = FALSE, warning = FALSE)
```
In the crime lecture, we saw how we can use heatmaps to give a 2-dimensional representation of 3-dimensional data: we made heatmaps of crime counts by time of the day and day of the week. In this problem, we'll learn how to use histograms to show counts by one variable, and then how to visualize 3 dimensions by creating multiple histograms.

We'll use the parole data parole.csv from Unit 3. Before, we used this data to predict parole violators. Now, let's try to get a little more insight into this dataset using histograms. As a reminder, the variables in this dataset are:

*male* = 1 if the parolee is male, 0 if female

*race* = 1 if the parolee is white, 2 otherwise

*age* = the parolee's age in years at the time of release from prison

*state* = a code for the parolee's state. 2 is Kentucky, 3 is Louisiana, 4 is Virginia, and 1 is any other state. These three states were selected due to having a high representation in the dataset.

*time.served* = the number of months the parolee served in prison (limited by the inclusion criteria to not exceed 6 months).

*max.sentence* = the maximum sentence length for all charges, in months (limited by the inclusion criteria to not exceed 18 months).

*multiple.offenses* = 1 if the parolee was incarcerated for multiple offenses, 0 otherwise.

*crime* = a code for the parolee's main crime leading to incarceration. 2 is larceny, 3 is drug-related crime, 4 is driving-related crime, and 1 is any other crime.

*violator* = 1 if the parolee violated the parole, and 0 if the parolee completed the parole without violation.

## Loading the Data

Using the read.csv function, load the dataset parole.csv and call it parole. Since male, state, and crime are all unordered factors, convert them to factor variables.
```{r}
parole <- read.csv("parole.csv")
parole$male <- as.factor(parole$male)
parole$state <- as.factor(parole$state)
parole$crime <- as.factor(parole$crime)
```
What fraction of parole violators are female?
```{r}
table(parole$male, parole$violator)
14/(14+64)
```
In this dataset, which crime is the most common in Kentucky?
```{r}
table(parole$state, parole$crime)
```
## Creating a Basic Histogram

Create a histogram to find out the distribution of the age of parolees.
```{r}
library(ggplot2)
ggplot(parole) + geom_histogram(aes(x=age), binwidth = 5, boundary = 0, color = "black", fill = "cornflowerblue")
```
```{r}
ggplot(parole) + geom_histogram(aes(x=age), binwidth = 5, boundary = 0, color = "blue", fill = "cornflowerblue")
```
## Adding Another Dimension

Now suppose we are interested in seeing how the age distribution of male parolees compares to the age distribution of female parolees.

One option would be to create a heatmap with age on one axis and male (a binary variable in our data set) on the other axis. Another option would be to stick with histograms, but to create a separate histogram for each gender. ggplot has the ability to do this automatically using the facet_grid command.
```{r}
ggplot(parole) + geom_histogram(aes(x=age), binwidth = 5) + facet_grid(male~.)
```
Now change the facet_grid argument to be ".~male" instead of "male~.". What does this do?
```{r}
ggplot(parole) + geom_histogram(aes(x=age), binwidth = 5) + facet_grid(.~male)
```

An alternative to faceting is to simply color the different groups differently. To color the data points by group, we need to tell ggplot that a property of the data (male or not male) should be translated to an aesthetic property of the histogram. We can do this by setting the fill parameter within the aesthetic to male.
```{r}
ggplot(parole) + geom_histogram(aes(x = age, fill = male), binwidth = 5)
```

Since we didn't specify colors to use, ggplot will use its default color selection. Let's change this by defining our own color palette.
```{r}
colorPalette = c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```
This is actually a colorblind-friendly palette, desribed on this Cookbook for R page. Now, generate your histogram again
```{r}
ggplot(data = parole, aes(x = age, fill = male)) + geom_histogram(binwidth = 5, boundary = 0) + scale_fill_manual(values=colorPalette)
```

Coloring the groups differently is a good way to see the breakdown of age by sex within the single, aggregated histogram. However, the bars here are stacked, meaning that the height of the orange bars in each age bin represents the total number of parolees in that age bin, not just the number of parolees in that group.

An alternative to a single, stacked histogram is to create two histograms and overlay them on top of each other. This is a simple adjustment to our previous command.
```{r}
ggplot(data = parole, aes(x = age, fill = male)) + geom_histogram(binwidth = 5, alpha = 0.5, position = "identity", boundary = 0) + scale_fill_manual(values=colorPalette)
```

## Time Served

Now let's explore another aspect of the data: the amount of time served by parolees. Create a basic histogram like the one we created in Problem 2, but this time with time.served on the x-axis. Set the bin width to one month.
```{r}
ggplot(parole) + geom_histogram(aes(x = time.served), binwidth = 1)
```

Change the binwidth to 0.1 months. Now what is the most common length of time served, according to the histogram?
```{r}
ggplot(parole) + geom_histogram(aes(x = time.served), binwidth = 0.1)
```

Be careful when choosing the binwidth - it can significantly affect the interpretation of a histogram! When visualizing histograms, it is always a good idea to vary the bin size in order to understand the data at various granularities.

Now, suppose we suspect that it is unlikely that each type of crime has the same distribution of time served. To visualize this, change the binwidth back to 1 month, and use facet_grid to create a separate histogram of time.served for each value of the variable crime.
```{r}
ggplot(parole) + geom_histogram(aes(x = time.served), binwidth = 1) + facet_grid(crime ~.)
```

Now, instead of faceting the histograms, overlay them. Remember to set the position and alpha parameters so that the histograms are not stacked. Also, make sure to indicate that the fill aesthetic should be "crime".
```{r}
ggplot(parole) + geom_histogram(aes(x = time.served, fill = crime), position = "identity", alpha = 0.5, binwidth = 1)
```
